---
title: "Impedance functions for Census Divisions"
author: "Bruno Santos & Antonio Paez"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
bibliography: skeleton.bib
link-citations: yes
editor_options: 
  markdown: 
    wrap: 72
---

```{r clean-workspace, include=FALSE}
# cleaning objects from the workspace 
rm(list = ls())
```

```{r setup, include=FALSE}
# layout configuration 
library(tufte) # not installed on this machine
library(knitr) # A General-Purpose Package for Dynamic Report Generation in R
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

# Introduction

This Rmarkdown file is part of the
[**CommuteCA**](https://github.com/dias-bruno/CommuteCA) package.
This package was created in conjunction with the office of the
[*Research Data Center* at *McMaster
University*](https://rdc.mcmaster.ca/), the [*Sherman Centre for Digital
Scholarship*](https://scds.ca/) and the [*Mobilizing
Justice*](https://mobilizingjustice.ca/)[^1].

[^1]: The Mobilizing Justice project is a multidisciplinary and
    multi-sector collaboration with the objective of understand and
    address transportation poverty in Canada and to improve the
    well-being of Canadians at risk of transport poverty. The Social
    Sciences and Humanities Research Council (SSRHC) has provided
    funding for the project, which was created by an unprecedented
    alliance of academics from various Canadian provinces and
    institutions, transportation firms, and nonprofit organizations

The main objective of the *CommuteCA* package is to develop
standardized methods for transport analysis in research, especially for
studies using Statistics Canada surveys. Among the available surveys, we
focused our efforts on the [*2021 Census of
Population*](https://www12.statcan.gc.ca/census-recensement/index-eng.cfm),
which contain valuable variables for transportation research.

This R markdown presents the methodology used to obtain the impedance functions for Canadian census divisions, available in *CommuteCA* package. Here, we record the method for obtaining `impedance_functions_pcd.rda` shared in the `CommuteCA/data/` folder.

## Impedance functions

Load the packages:

```{r load-packages}
library(dplyr)# A Grammar of Data Manipulation 
library(fitdistrplus) # Help to Fit of a Parametric Distribution to Non-Censored or Censored Data
library(scales) # Scale data column-wise in a computationally efficient way
library(here) # enable easy file referencing in project-oriented workflows
```

Reading census data and creating a R data frame:

```{r}
files_address <- paste0(here(),
                        "/data-ignore-inputs/census_test_v3.csv") # Census address
```


```{r load-census-data}
census <- read.csv(files_address, header = TRUE)
```

| ⚠️**NOTE:**  If the code above did not run correctly, you probably are experiencing a file address error. Try to identify the correct address and update the chunk named `census-file-address` to continue.

The original census dataset has a wide variety of attributes. However,
we will only work with a selection of these variables. The next chunk
selects only the variables previously chosen for our analysis, reducing
the size of the dataframe and making data processing faster:

```{r select-variables-census}
census <- census %>% 
          dplyr::select("PRCDDA",
                 "Pr",
                 "CMA",
                 "PCD",
                 "CompW1",
                 "LBR_FORC",
                 "PWDUR",
                 "PwMode")
```

Renaming the variable `PwMode`:

```{r rename-PwMode}
census <- census %>% 
  mutate(PwMode = 
           case_when(PwMode > 0 & PwMode <= 2 ~ "Car/motor",
                     PwMode == 9 ~ "Car/motor",
                     PwMode >= 3 & PwMode <= 6  ~ "Transit",
                     PwMode == 7  ~ "Walk",
                     PwMode == 8  ~ "Bike"),
         PwMode = factor(PwMode, levels = c("Bike", "Walk", "Car/motor", "Transit")),
         CMA_type = 
           case_when(CMA == 0 ~ 'Territories (outside CAs)',
                     CMA > 0 & CMA < 996 ~ 'CMA/CA',
                     CMA == 996 ~ 'Strong metropolitan influenced zone',
                     CMA == 997 ~ 'Moderate metropolitan influenced zone',
                     CMA == 998 ~ 'Weak metropolitan influenced zone',
                     CMA == 999 ~ 'No metropolitan influenced zone'),
          CMA_type = factor(CMA_type, levels = c("Territories (outside CAs)",
                                                 "CMA/CA", 
                                                 "Strong metropolitan influenced zone", 
                                                 "Moderate metropolitan influenced zone",
                                                 'Weak metropolitan influenced zone',
                                                 'No metropolitan influenced zone')),
         Pr = case_when(
    Pr == 10 ~ "Newfoundland and Labrador",
    Pr == 11 ~ "Prince Edward Island",
    Pr == 12 ~ "Nova Scotia",
    Pr == 13 ~ "New Brunswick",
    Pr == 24 ~ "Quebec",
    Pr == 35 ~ "Ontario",
    Pr == 46 ~ "Manitoba",
    Pr == 47 ~ "Saskatchewan",
    Pr == 48 ~ "Alberta",
    Pr == 59 ~ "British Columbia",
    Pr == 60 ~ "Yukon",
    Pr == 61 ~ "Northwest Territories",
    Pr == 62 ~ "Nunavut",
    TRUE ~ as.factor(Pr)))
```

According to the Census 2021, an 'Employed' person refers to those who,
during the reference period, had a labour force status of
'employed.'[^11]. In addition, the labour force is made up of the
employed population and the unemployed population. This variable can
assume the values:

[^11]: Those who, during the reference period: (a) Did any work at all
    at a job or business, that is, paid work in the context of an
    employer-employee relationship, or self-employment. This also
    includes persons who did unpaid family work, which is defined as
    unpaid work contributing directly to the operation of a farm,
    business or professional practice owned and operated by a related
    member of the same household; or (b) Had a job but were not at work
    due to factors such as their own illness or disability, personal or
    family responsibilities, vacation or a labour dispute. This category
    excludes persons not at work because they were on layoff or between
    casual jobs, and those who did not then have a job (even if they had
    a job to start at a future date).

-   -3: Not Applicable, \< 15 years
-   1: In Labour Force, Employed
-   2: In Labour Force, Unemployed
-   3: Not in Labour Force

Each employee also declares in which DA their workplace is located. This
information is important for the accessibility analysis, as it allows us
to count the number of job opportunities in each DA.

```{r employed-selection}
# Employed respondents
census_employed <- census %>% 
                   filter(LBR_FORC == 1)
```

Now that we already have the travel times for each trip, we will try to create a impedance function that better
describes the travel pattern of the respondents in each destination. To do so,
we'll first create a function that based on the lowest Akaike
Information Criterion (AIC)[^12], selects the better function between
the distributions:
[exponential](https://en.wikipedia.org/wiki/Exponential_distribution),
[gamma](https://en.wikipedia.org/wiki/Gamma_distribution),
[log-normal](https://en.wikipedia.org/wiki/Log-normal_distribution),
[normal](https://en.wikipedia.org/wiki/Normal_distribution), and
[uniform](https://en.wikipedia.org/wiki/Continuous_uniform_distribution).

[^12]: The Akaike Information Criterion (AIC) estimates prediction error
    and model quality for a given data set.

We'll use the
[{fitdistrplus}](https://cloud.r-project.org/web/packages/fitdistrplus/index.html)
to determine the parameters that best fit the TLD. The Moment Matching Estimation (MME) fitting method and the Nelder-Mead
direct optimization algorithm are used [@mullerdutang2015].

```{r lowest-aic-function}
# Select the distribution function based on the lowest AIC value
lowest_aic <- function(values){
  
  min_aic <- min(values)
  
  if(!is.na(lnorm_$aic) & min_aic == lnorm_$aic){
    choosen_f <- lnorm_
  }
  
  else if  (!is.na(gamma_$aic) & min_aic == gamma_$aic){
    choosen_f <- gamma_
  }  
  
  else if  (!is.na(unif_$aic) & min_aic == unif_$aic){
    choosen_f <- unif_
  }  
  
  else if  (!is.na(norm_$aic) & min_aic == norm_$aic){
    choosen_f <- norm_
  }
  else if  (!is.na(exp_$aic) & min_aic == exp_$aic){
    choosen_f <- exp_
  }
  
  return(choosen_f)
}

# Test the distributions
test_distributions_weighted <- function(x, weights){
  gamma_ <<- fitdistrplus::fitdist(data=x, "gamma", method="mme", weights = weights)
  lnorm_ <<- fitdistrplus::fitdist(data=x, "lnorm", method="mme", weights = weights)
  norm_ <<- fitdistrplus::fitdist(data=x, "norm", method="mme", weights = weights)
  exp_ <<- fitdistrplus::fitdist(data=x, "exp", method="mme", weights = weights)
  unif_ <<- fitdistrplus::fitdist(data=x, "unif", method="mme", weights = weights)
  
  values <- c(lnorm_$aic, gamma_$aic, unif_$aic, norm_$aic, exp_$aic)
  values <- values[!is.na(values)]
  
  chosen_function <- lowest_aic(values)

  return(chosen_function)
}
```

for PCD: 

```{r}
pcd_impedance_functions <- data.frame()

for(division in unique(census_employed$PCD)){
  
  for(mode in unique(census_employed[census_employed$PCD == division & !is.na(census_employed$PwMode), ]$PwMode)){
      
       census_subset <-  census_employed %>%
         filter(PCD == division, PwMode == mode, PWDUR > 0)
        
        if(nrow(census_subset) > 1){
          # print(paste("Calculating the best function for the census subdivision ", 
          #           division, " and ",  
          #           mode,
          #           " transportaion mode."))
  
          assign(paste0("f_",division, "_", mode),
               test_distributions_weighted(
               x = census_subset$PWDUR,
               weights = round(census_subset$CompW1)))
          
         current_function <- get(paste0("f_",division, "_", mode))
         
         # lnorm impedance 
        if(current_function$distname == "lnorm"){
             fname <-  current_function$distname
             estimate_1 <- current_function$estimate[1]
             estimate_2 <- current_function$estimate[2]
            }
            
            # Unif impedance 
            else if(current_function$distname == "unif") {
             fname <-  current_function$distname
             estimate_1 <- 0
             estimate_2 <- current_function$estimate[2]
             
            # Exponential impedance 
            } else if(current_function$distname == "exp") {
             fname <-  current_function$distname
             estimate_1 <- current_function$estimate
             estimate_2 <- 0 
             
                # Gamma impedance 
            } else if (current_function$distname == "gamma"){  
             fname <-  current_function$distname
             estimate_1 <- current_function$estimate[1]
             estimate_2 <- current_function$estimate[2]
             
            # Norm impedance 
            } else if (current_function$distname == "norm"){
             fname <-  current_function$distname
             estimate_1 <- current_function$estimate[1]
             estimate_2 <- current_function$estimate[2]
            }
            
        df_f <- data.frame(
          "PCD" = division, 
          "Pr" = census_subset$Pr[1],
          "CMA" = census_subset$CMA[1],
          "PwMode" = mode, 
          "distribution" = fname,
          "est_1" = estimate_1,
          "est_2" = estimate_2,
          "loglike"  = current_function$loglik,
          "AIC"  = current_function$aic,
          "BIC"  = current_function$bic,
          "n" = current_function$n,
          "Weighted_sum" = sum(current_function$weights))
        
        pcd_impedance_functions <- rbind(pcd_impedance_functions, df_f)
        
        rownames(pcd_impedance_functions) <- NULL
        
        }
    }
}
```

for CMA: 

```{r}
cma_impedance_functions <- data.frame()

for(division in unique(census_employed$CMA)){
  
  for(mode in unique(census_employed[census_employed$CMA == division & !is.na(census_employed$PwMode), ]$PwMode)){
      
       census_subset <-  census_employed %>%
         filter(CMA == division, PwMode == mode, PWDUR > 0)
        
        if(nrow(census_subset) > 1){
          # print(paste("Calculating the best function for the CMA ", 
          #           division, " and ",  
          #           mode,
          #           " transportaion mode."))
  
          assign(paste0("f_",division, "_", mode),
               test_distributions_weighted(
               x = census_subset$PWDUR,
               weights = round(census_subset$CompW1)))
          
         current_function <- get(paste0("f_",division, "_", mode))
         
         # lnorm impedance 
        if(current_function$distname == "lnorm"){
             fname <-  current_function$distname
             estimate_1 <- current_function$estimate[1]
             estimate_2 <- current_function$estimate[2]
            }
            
            # Unif impedance 
            else if(current_function$distname == "unif") {
             fname <-  current_function$distname
             estimate_1 <- 0
             estimate_2 <- current_function$estimate[2]
             
            # Exponential impedance 
            } else if(current_function$distname == "exp") {
             fname <-  current_function$distname
             estimate_1 <- current_function$estimate
             estimate_2 <- 0 
             
                # Gamma impedance 
            } else if (current_function$distname == "gamma"){  
             fname <-  current_function$distname
             estimate_1 <- current_function$estimate[1]
             estimate_2 <- current_function$estimate[2]
             
            # Norm impedance 
            } else if (current_function$distname == "norm"){
             fname <-  current_function$distname
             estimate_1 <- current_function$estimate[1]
             estimate_2 <- current_function$estimate[2]
            }
            
        df_f <- data.frame(
          "CMA" = division, 
          "Pr" = census_subset$Pr[1],
          "PwMode" = mode, 
          "distribution" = fname,
          "est_1" = estimate_1,
          "est_2" = estimate_2,
          "loglike"  = current_function$loglik,
          "AIC"  = current_function$aic,
          "BIC"  = current_function$bic,
          "n" = current_function$n,
          "Weighted_sum" = sum(current_function$weights))
        
        cma_impedance_functions <- rbind(cma_impedance_functions, df_f)
        
        rownames(cma_impedance_functions) <- NULL
        
        }
    }
}
```


for Province: 

```{r}
pr_impedance_functions <- data.frame()

for(division in unique(census_employed$Pr)){
  
  for(type in unique(census_employed[census_employed$Pr == division, ]$CMA_type)){
  
  for(mode in unique(census_employed[census_employed$CMA_type == type &
                                     census_employed$Pr == division &
                                     !is.na(census_employed$PwMode), ]$PwMode)){
      
       census_subset <-  census_employed %>%
         filter(Pr == division, PwMode == mode, PWDUR > 0, CMA_type == type)
        
        if(nrow(census_subset) > 1){
          print(paste("Calculating the best function for the Province ", 
                    division, ", considering ", type, " and ",   
                    mode, " transportaion mode."))
  
          assign(paste0("f_",division, "_", mode),
               test_distributions_weighted(
               x = census_subset$PWDUR,
               weights = round(census_subset$CompW1)))
          
         current_function <- get(paste0("f_",division, "_", mode))
         
         # lnorm impedance 
        if(current_function$distname == "lnorm"){
             fname <-  current_function$distname
             estimate_1 <- current_function$estimate[1]
             estimate_2 <- current_function$estimate[2]
            }
            
            # Unif impedance 
            else if(current_function$distname == "unif") {
             fname <-  current_function$distname
             estimate_1 <- 0
             estimate_2 <- current_function$estimate[2]
             
            # Exponential impedance 
            } else if(current_function$distname == "exp") {
             fname <-  current_function$distname
             estimate_1 <- current_function$estimate
             estimate_2 <- 0 
             
                # Gamma impedance 
            } else if (current_function$distname == "gamma"){  
             fname <-  current_function$distname
             estimate_1 <- current_function$estimate[1]
             estimate_2 <- current_function$estimate[2]
             
            # Norm impedance 
            } else if (current_function$distname == "norm"){
             fname <-  current_function$distname
             estimate_1 <- current_function$estimate[1]
             estimate_2 <- current_function$estimate[2]
            }
            
        df_f <- data.frame(
          "Pr" = division,
          "CMA_type" = type,
          "PwMode" = mode, 
          "distribution" = fname,
          "est_1" = estimate_1,
          "est_2" = estimate_2,
          "loglike"  = current_function$loglik,
          "AIC"  = current_function$aic,
          "BIC"  = current_function$bic,
          "n" = current_function$n,
          "Weighted_sum" = sum(current_function$weights))
        
        pr_impedance_functions <- rbind(pr_impedance_functions, df_f)
        
        rownames(pr_impedance_functions) <- NULL
        
        }
  }
}
}
```

## Confidentiality vetting

Confidentiality vetting is the process of reviewing the results to be released by the Research Data Center to ensure that confidentiality risks for StatCan respondents are minimized. 

The following rules apply to our results:

- Statistics must not be released for identifiable areas with less than 40 persons ($\sum{CompW1} ≥ 40$).
- Statistics must be based on at least 5 respondents ($n > 5$), with a sum of weights of at least 10 ($\sum{CompW1} > 10$), to be releasable. 

The block above checks the total population of each geographic unit:

```{r da-vetting}
vetting_pcd <- census %>%
  group_by(PCD) %>%
  summarise(pop_area = sum(CompW1))

vetting_cma <- census %>%
  group_by(CMA) %>%
  summarise(pop_area = sum(CompW1))

vetting_pr <- census %>%
  group_by(Pr) %>%
  summarise(pop_area = sum(CompW1))

pcd_impedance_functions <- pcd_impedance_functions %>% 
  left_join(vetting_pcd, by = 'PCD') 

cma_impedance_functions <- cma_impedance_functions %>% 
  left_join(vetting_cma, by = 'CMA') 

pr_impedance_functions <- pr_impedance_functions %>% 
  left_join(vetting_pr, by = 'Pr') 
```

Creating new files with the total unit population:

```{r creating-release}
pcd_impedance_functions <- pcd_impedance_functions %>% 
  filter(n > 4, Weighted_sum > 9, pop_area > 40)

cma_impedance_functions <- cma_impedance_functions %>% 
  filter(n > 4, Weighted_sum > 9, pop_area > 40)

pr_impedance_functions <- pr_impedance_functions %>% 
  filter(n > 4, Weighted_sum > 9, pop_area > 40)
```


# Data export

To finalize the methodology of this R markdown, we will export the
processed data. 

Creating the directory:

```{r directory-export}
diretorio_export <- paste0(here::here(),"/data-ignore-inputs/") #Update the address to export the data, if necessary
dir.create(diretorio_export, recursive = TRUE)
```

Exporting tables:

```{r export-land-use-release}
# PCD
write.csv(pcd_impedance_functions %>% dplyr::select(-pop_area), paste0(diretorio_export, "pcd_impedance_functions_release.csv"), row.names=FALSE)

write.csv(pcd_impedance_functions, paste0(diretorio_export, "pcd_impedance_functions_support.csv"), row.names=FALSE)

# CMA
write.csv(cma_impedance_functions %>% dplyr::select(-pop_area), paste0(diretorio_export, "cma_impedance_functions_release.csv"), row.names=FALSE)

write.csv(cma_impedance_functions, paste0(diretorio_export, "cma_impedance_functions_support.csv"), row.names=FALSE)

# PR 
write.csv(pr_impedance_functions %>% dplyr::select(-pop_area), paste0(diretorio_export, "pr_impedance_functions_release.csv"), row.names=FALSE)

write.csv(pr_impedance_functions, paste0(diretorio_export, "pr_impedance_functions_support.csv"), row.names=FALSE)
```

