---
title: "Visualizing accessibility measures (_Dissemination area_ level)"
author: "Bruno Santos & Antonio Paez"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
bibliography: skeleton.bib
link-citations: yes
editor_options: 
  markdown: 
    wrap: 72
---

```{r clean-workspace, include=FALSE}
# cleaning objects from the workspace 
rm(list = ls())
```

```{r setup, include=FALSE}
# layout configuration 
library(tufte)
library(knitr)
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

# Introduction

This Rmarkdown file is part of the
[**CommuteCA**](https://github.com/dias-bruno/CommuteCA) package.
This package was created in conjunction with the office of the
[*Research Data Center* at *McMaster
University*](https://rdc.mcmaster.ca/), the [*Sherman Centre for Digital
Scholarship*](https://scds.ca/) and the [*Mobilizing
Justice*](https://mobilizingjustice.ca/)[^1].

[^1]: The Mobilizing Justice project is a multidisciplinary and
    multi-sector collaboration with the objective of understand and
    address transportation poverty in Canada and to improve the
    well-being of Canadians at risk of transport poverty. The Social
    Sciences and Humanities Research Council (SSRHC) has provided
    funding for the project, which was created by an unprecedented
    alliance of academics from various Canadian provinces and
    institutions, transportation firms, and nonprofit organizations

The [**CommuteCA**](https://github.com/dias-bruno/CommuteCA) R
package was created to develop standardized methods for transport
analysis in research, particularly for analysis using the [*2021 Census of
Population*](https://www12.statcan.gc.ca/census-recensement/index-eng.cfm) from Statistics Canada. We focused our efforts on the [_Commuting Reference Guide_](https://www12.statcan.gc.ca/census-recensement/2021/ref/98-500/011/98-500-x2021011-eng.cfm),
which provides valuable variables and information on commuting for the Canadian population aged 15 and older living in private households. 

After obtaining the accessibility measures for our study area, in this
`R` Markdown we will visualize the measure in form of maps.

## Suggested Readings

-   Soukhov, Anastasia, Antonio Paez, Christopher D Higgins, and Moataz
    Mohamed. 2023."Introducing Spatial Availability, a
    Singly-Constrained Measure of Competitive Accessibility." *Plos One*
    18 (1): e0278468.
-   Soukhov, A., Páez, A. (2024). Accessibility analysis for planning
    applications (Report No. MJ-A2-0002). Mobilizing Justice.
    <https://github.com/soukhova/MJ-Accessibility-Blogs>
-   Pereira, Rafael H. M.; Herszenhut, Daniel. Introduction to urban
    accessibility: a practical guide with R. Rio de Janeiro: Ipea, 2023.
    152 p. ISBN: 978-65-5635-065-3. DOI:
    <http://dx.doi.org/9786556350653>.
-   Soukhov A, Tarriño-Ortiz J, Soria-Lara JA, Páez A (2024) Multimodal
    spatial availability: A singly-constrained measure of accessibility
    considering multiple modes. PLoS ONE 19(2): e0299077.
    <https://doi.org/10.1371/journal.pone.0299077>

## Data

The data used in this R markdown is the output of the previous `R`
markdown, in which we calculated accessibility and spatial availability
for our study area. The methodology described here is applicable to the
results obtained through the synthetic or original census survey.
Additionally, we will use the dissemination areas (DA) geometry,
available in this package, to plot maps of the study area.

# Let's code!

Load the packages:

```{r load-packages}
library(CommuteCA) 
library(dplyr)# A Grammar of Data Manipulation 
library(ggplot2) # Create Elegant Data Visualizations Using the Grammar of Graphics
library(RColorBrewer) # color schemes for maps (and other graphics) designed
library(scales) # for scaling data
library(sf) # support for simple features, a standardized way to encode spatial vector data
library(tmap) # thematic maps
library(tidyr) # tidying data
```

Reading the tables:

```{r}
output_folder <- paste0(here::here(),"/data-raw/DA/output/PCD3520/")
```


```{r reading-tables}
HT_mode <- read.csv(paste0(output_folder, "HT_mode_original.csv"))
HT <- read.csv(paste0(output_folder, "HT_general_original.csv"))
SA_mode <- read.csv(paste0(output_folder, "SA_mode_original.csv"))
SA <- read.csv(paste0(output_folder, "SA_general_original.csv"))
land_use_mode <- read.csv(paste0(output_folder, "land_use_mode_original_DA.csv"))
land_use_general <- read.csv(paste0(output_folder, "land_use_general_original_DA.csv"))
```

Reading the DA geometries:
```{r}
data("dissemination_areas_ca21.rda")
data("census_divisions_ca21.rda")
```

Selecting only the DA in our study area:

```{r}
DA <- dissemination_areas_ca21

DA <- DA %>% 
  filter(PCD %in% (unique(substr(unique(land_use_general$PRCDDA), 1,4)))) # It will select the dissemination areas with PCD codes within our study area

CD <- census_divisions_ca21 

CD <- CD %>% 
  filter(CDUID %in% (unique(DA$PCD))) # It will select the census division of our study area
```

Merging the DA with the accessibility measures tables:

```{r}
names(DA)[1] <- "PRCDDA"

mode_metrics_da <- 
  expand.grid(PRCDDA = unique(DA$PRCDDA), PwMode = unique(HT_mode$PwMode)) %>%
  full_join(DA, by = "PRCDDA") %>%
  mutate(PRCDDA = as.numeric(PRCDDA)) %>%
  full_join(HT_mode, by = c("PRCDDA", "PwMode")) %>%
  full_join(SA_mode, by = c("PRCDDA", "PwMode")) %>% 
  full_join(land_use_mode[,c("PRCDDA","PwMode","labour_force")], by = c("PRCDDA", "PwMode"))

mode_metrics_da[is.na(mode_metrics_da$HT_im),]$HT_im <- 0
mode_metrics_da[is.na(mode_metrics_da$SA_im),]$SA_im <- 0
mode_metrics_da[is.na(mode_metrics_da$labour_force),]$labour_force <- 0

mode_metrics_da_sf <- st_as_sf(mode_metrics_da, crs = st_crs(DA))
```

Preparing the tables to plot the maps:

```{r}
# Hansen
HT_metrics_da <- DA %>% 
  mutate(PRCDDA = as.numeric(PRCDDA)) %>%
  full_join(HT, by = "PRCDDA") %>%
  full_join(land_use_general, by ="PRCDDA")

HT_metrics_da[is.na(HT_metrics_da$HT_i),]$HT_i <- 0
HT_metrics_da[is.na(HT_metrics_da$labour_force),]$labour_force <- 0
HT_metrics_da[is.na(HT_metrics_da$jobs),]$jobs <- 0

# Spatial Availability
SA_metrics_da <- DA %>% 
  mutate(PRCDDA = as.numeric(PRCDDA)) %>%
  full_join(SA, by = "PRCDDA") %>% 
  full_join(land_use_general, by ="PRCDDA")

SA_metrics_da[is.na(SA_metrics_da$SA_i),]$SA_i <- 0
SA_metrics_da[is.na(SA_metrics_da$labour_force),]$labour_force <- 0
SA_metrics_da[is.na(SA_metrics_da$jobs),]$jobs <- 0
```

## Visualizing the accessibility measures with maps

Plotting the count of people within the labour force and count of job
opportunities in each DA:

```{r}
labour_force_plot <- tm_shape(HT_metrics_da) + 
   tm_polygons("labour_force",
              style = "cont",
              palette = "Reds",
              title = " ",
              border.col = NULL) + 
  tm_shape(CD) +
  tm_borders("black", lwd=0.5) +
  tm_scale_bar(position = c("right", "bottom"))+
  tm_compass(position = c("left", "top"), size=1.0)  + 
  tm_layout(legend.outside = FALSE,
            legend.position = c("left", "bottom"),
            panel.labels = "Labour force")

jobs_plot <- tm_shape(SA_metrics_da) + 
   tm_polygons("jobs",
              style = "cont",
              palette = "Blues",
              title = " ",
              border.col = NULL) + 
  tm_shape(CD) +
  tm_borders("black", lwd=0.5) + 
  tm_scale_bar(position = c("right", "bottom"))+
  tm_compass(position = c("left", "top"), size=1.0)  + 
  tm_layout(legend.outside = FALSE,
            legend.position = c("left", "bottom"),
            panel.labels = "Job opportunities")

lf_jobs_DA <- tmap_arrange(labour_force_plot, jobs_plot, ncol = 2)

tmap_save(lf_jobs_DA, 
          paste0(output_folder,"lf_jobs_DA.jpg"), 
          width = 20, 
          height = 15, 
          units = "cm", 
          dpi = 300)

lf_jobs_DA
```

Plotting the accessibility measures:

```{r}
ht_plot <- tm_shape(HT_metrics_da) + 
   tm_polygons("HT_i",
              style = "cont",
              palette = "Purples",
              title = " ",
              border.col = NULL) + 
  tm_shape(CD) +
  tm_borders("black", lwd=0.5) + 
  tm_scale_bar(position = c("right", "bottom"))+
  tm_compass(position = c("left", "top"), size=1.0)  + 
  tm_layout(legend.outside = FALSE,
            legend.position = c("left", "bottom"),
            panel.labels = "Hansen-type accessibility of jobs")

sa_plot <- tm_shape(SA_metrics_da) + 
   tm_polygons("SA_i",
              style = "cont",
              palette = "Oranges",
              title = " ",
              border.col = NULL) + 
  tm_shape(CD) +
  tm_borders("black", lwd=0.5) +
  tm_scale_bar(position = c("right", "bottom"))+
  tm_compass(position = c("left", "top"), size=1.0)  + 
  tm_layout(legend.outside = FALSE,
            legend.position = c("left", "bottom"),
            panel.labels = "Spatial availability of jobs")

HT_SA_DA <- tmap_arrange(ht_plot, sa_plot, ncol = 2)

tmap_save(HT_SA_DA, 
          paste0(output_folder,"HT_SA_DA.jpg"), 
          width = 20, 
          height = 15, 
          units = "cm", 
          dpi = 300)

HT_SA_DA
```

Plotting the jobs per labour force, and the spatial availability per
labour force:

```{r}
HT_metrics_da$jobs_per_lf <- HT_metrics_da$jobs / HT_metrics_da$labour_force

HT_metrics_da <- HT_metrics_da %>%
  mutate(jobs_per_lf = case_when(
    jobs == 0 & labour_force == 0 ~ 0,
    labour_force == 0 & jobs != 0 ~ jobs,  
    TRUE ~ jobs_per_lf))

jobs_per_labour_force <- tm_shape(HT_metrics_da) + 
   tm_polygons("jobs_per_lf",
              style = "fixed",
              breaks = c(0, 0.5, 1, 2.5, 5, 10, Inf),
              palette = "RdYlGn",
              title = " ",
              border.col = NULL) + 
  tm_shape(CD) +
  tm_borders("black", lwd=0.5) + 
  tm_scale_bar(position = c("right", "bottom"))+
  tm_compass(position = c("left", "top"), size=1.0)  + 
  tm_layout(legend.outside = FALSE,
            legend.position = c("left", "bottom"),
            panel.labels = "Jobs per person in labour fource")

SA_metrics_da$SA_per_lf <- SA_metrics_da$jobs / SA_metrics_da$labour_force

SA_metrics_da <- SA_metrics_da %>%
  mutate(SA_per_lf = case_when(
    jobs == 0 & labour_force == 0 ~ 0,
    labour_force == 0 & jobs != 0 ~ jobs,  
    TRUE ~ SA_per_lf))

SA_per_labour_force <- tm_shape(SA_metrics_da) + 
   tm_polygons("SA_per_lf",
              style = "fixed",
              breaks = c(0, 0.5, 1, 2.5, 5, 10, Inf),
              palette = "RdYlGn",
              title = " ",
              border.col = NULL) + 
  tm_shape(CD) +
  tm_borders("black", lwd=0.5) + 
  tm_scale_bar(position = c("right", "bottom"))+
  tm_compass(position = c("left", "top"), size=1.0)  + 
  tm_layout(legend.outside = FALSE,
            legend.position = c("left", "bottom"),
            panel.labels = "Spatial availability per person in labour force")

jobs_SA_per_lf_DA <- tmap_arrange(jobs_per_labour_force, SA_per_labour_force, ncol = 2)


tmap_save(jobs_SA_per_lf_DA, 
          paste0(output_folder,"jobs_SA_per_lf_DA.jpg"), 
          width = 20, 
          height = 15, 
          units = "cm", 
          dpi = 300)

jobs_SA_per_lf_DA
```

## Visualizing the accessibility measures by transportation modes with maps

Visualizing the total labour force by transportation mode:

```{r}
total_pop_by_mode_DA <- tm_shape(mode_metrics_da_sf) + 
   tm_polygons("labour_force",
              style = "cont",
              palette = "Reds",
              title = " ",
              border.col = NULL) +  
   tm_facets(by = "PwMode") +       
   tm_scale_bar(position = c("right", "bottom")) +
   tm_compass(position = c("left", "top"), size = 1.0) + 
   tm_shape(CD) +
   tm_borders("black", lwd=0.5)

tmap_save(total_pop_by_mode_DA, 
          paste0(output_folder,"total_pop_by_mode_DA.jpg"), 
          width = 20, 
          height = 20, 
          units = "cm", 
          dpi = 300)

total_pop_by_mode_DA
```

Visualizing the Hansen-type accessibility by transportation mode:

```{r}
HT_by_mode_DA <- tm_shape(mode_metrics_da_sf) + 
   tm_polygons("HT_im",
              style = "cont",
              palette = "Purples",
              title = " ",
              border.col = NULL) + 
  tm_scale_bar(position = c("right", "bottom"))+
  tm_compass(position = c("left", "top"), size=1.0)  + 
  tm_facets(by="PwMode") + 
   tm_shape(CD) +
   tm_borders("black", lwd=0.5)

tmap_save(HT_by_mode_DA, 
          paste0(output_folder,"HT_by_mode_DA.jpg"), 
          width = 20, 
          height = 20, 
          units = "cm", 
          dpi = 300)

HT_by_mode_DA
```

Visualizing the Spatial Availability by transportation mode:

```{r}
SA_by_mode_DA <- tm_shape(mode_metrics_da_sf) + 
   tm_polygons("SA_im",
              style = "cont",
              palette = "Oranges",
              title = " ",
              border.col = NULL) + 
  tm_scale_bar(position = c("right", "bottom"))+
  tm_compass(position = c("left", "top"), size=1.0)  + 
  tm_facets(by="PwMode") + 
   tm_shape(CD) +
   tm_borders("black", lwd=0.5)

tmap_save(SA_by_mode_DA, 
          paste0(output_folder,"SA_by_mode_DA.jpg"), 
          width = 20, 
          height = 20, 
          units = "cm", 
          dpi = 300)

SA_by_mode_DA
```

Computing of the labour force by transportation mode:

```{r}
total_pop <- sum(land_use_mode$labour_force)
total_SA <- sum(mode_metrics_da$SA_im)

modal_SA_percents <- mode_metrics_da %>% 
  group_by(PwMode) %>%
  summarize(total_labour_fource = sum(labour_force),
            SA_meam = mean(SA_im),
            SA_sum = sum(SA_im),
            perc_SA = sum(SA_im)/total_SA, 
            perc_pop= sum(labour_force)/total_pop)

modal_SA_long <- modal_SA_percents %>%
  pivot_longer(cols = c(perc_SA, perc_pop), names_to = "perc", values_to = "measure")
```

Visualizing the share of labour force and spatial availability by
transportation mode:

```{r}
Pop_SA_by_mode_DA <- ggplot(modal_SA_long, 
      aes(x = perc, y = measure, fill = PwMode, label = measure %>% percent(accuracy = 0.1))) +
      scale_x_discrete(labels = c(
      "perc_SA" = paste("Spatial Availability\n by Mode\n(Total: ", round(total_SA), ")"),
      "perc_pop" = paste("Population\n by Mode\n(Total: ", round(total_pop), ")") )) +
  geom_col() + 
  scale_fill_manual(values = c("#96CCD9","#49A695", "#A67360", "#8C3926")) + 
  geom_text(position = position_stack(vjust=0.5), color="black") +
  theme_classic() +
  theme(axis.ticks = element_blank(),
        axis.title = element_blank(),
        axis.line = element_blank(),
        axis.text.y = element_blank(),
        axis.text = element_text(color="black"),
        legend.position = "bottom",
        legend.title = element_blank())


ggsave(file = paste0(output_folder,"Pop_SA_by_mode_DA.jpg"), 
       plot = Pop_SA_by_mode_DA,
       width = 10, 
       height = 10, 
       units = "in", 
       dpi = 300)

Pop_SA_by_mode_DA
```
